---
title: "Lab 4: Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1- Answer questions on M3/A3
2- Answer questions on M4
3- Practice wrangling datasets with dplyr functions

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: 

```{r workspace setup, message = FALSE}
#Install packages
library(tidyverse)
library(lubridate)
library(here) #The here package allows for better control of relative paths

#Ensure that "here" points to your project folder
here()

#Read in the data
NTL.phys.data <- read.csv(
  file=here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE
)

#Show the datatype of the 'sampledate' column
str(NTL.phys.data$sampledate)

#Alternatively, use the tidyverse/dplyr "glimpse" function
glimpse(NTL.phys.data$sampledate)

# Change sampledate values into date objects
NTL.phys.data$sampledate <- mdy(NTL.phys.data$sampledate)
```

---

## Filter

Filtering allows us to choose certain rows (observations) in our dataset.

```{r filter data}
# note the data types of these two columns
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# dplyr filtering
NTL.phys.data.surface <- filter(NTL.phys.data, depth == 0)

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename)
NTL.phys.data.PeterPaul <- 
  filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake")) 
# filter(df, colname %in% c("factor_value"))
# %in% => select more than 1 factors
# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober <- filter(NTL.phys.data, daynum %in% c(152:304))
```

```{r Exercise 1 & 2}
# Exercise 1: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)
NTL.phys.data.1999 <- filter(NTL.phys.data, year4 == 1999)

# Exercise 2: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.
NTL.phys.data.Tue1990s <- filter(NTL.phys.data,lakename == "Tuesday Lake", year4 %in% c(1990:1999))

```
Question: Why don't we filter using row numbers?

> Answer: 

---

## Pipes

Pipe is another method to wrangle datasets that looks cleaner and is easier to read.  We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r Exercise 3 & 4}
#Example using pipes to wrangle data # %>% => the pipeline symbol => "Then"
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)

#Exercise 3: Using a pipe filter NTL.phys.data for Tuesday Lake from 1990 
# through 1999 only for July.
NTL.phys.data.Tue1990sJul <- NTL.phys.data %>% 
  filter(lakename == "Tuesday Lake", year4 %in% c(1990:1999), month(sampledate)== 7)

#Exercise 4: Using the data from part 3, a pipe, and the summarize() function 
# find the mean surface temperature (hints: you will need to add another
# filter for depth==0). Make sure you eliminate NAs before computing the means.
Tue1990sJul.surface <- NTL.phys.data.Tue1990sJul %>% 
  filter(depth==0) %>% 
  drop_na(temperature_C) %>% 
  summarise(meanT = mean(temperature_C))
  
  
```

## Gather and Spread

For gather we will use `pivot_longer` and for spread we will use `pivot_wider`.

```{r Exercise 5 & 6}
#Exercise 5: Gather irradiance data (measured in the water column and measured
#  on the deck of the sampling boat) into one column using pivot_longer. Name
#  the new column holding the irradiance type as "Irradiance_Type", and name the
#  new column holding the irradiance values as "Irradiance_Value".

NTL.pivot.longer <- NTL.phys.data %>% 
  pivot_longer(
    cols = c(irradianceWater, irradianceDeck),
    names_to = "Irradiance_Type",
    values_to = "Irradiance_Value"
  )
#Exercise 6: Spread temperatureC into more than one column based on the depth.

NTL.pivot.wider <- NTL.phys.data %>% 
  pivot_wider(
    names_from = depth,
    values_from = temperature_C
  )
```

