---
title: "Assignment 2: Coding Basics"
author: "Yin-Chia Yang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Generating a sequence and naming it
seq(1 ,30, 3)
num1 <- seq(1 ,30, 3)

#2. Computing the mean and mesian of the sequence
mean(num1)
mean_num1 <- mean(num1)

median(num1)
median_num1 <- median(num1)

#3. Comparing the mean and the median
if (mean_num1 > median_num1){
  print ("The mean is greater than the median.")
} else {
  print("The median is greater than the mean.")
} 

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
student_name <- c("Anna", "Bellie", "Cindy", "Diane" ) #character
test_score <- c(70, 65, 45, 55) #numeric
pass_test <- test_score >=50 #logical

df_test_result <- data.frame(
  Name = student_name,
  Test_Score = test_score,
  Passed = pass_test
)
print(df_test_result)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrices can only contain a single class of data, while data frames can consist of many different classes of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

```{r}
#if, else method1
pass_fail <- function(x){
  if (x > 49){
    print(TRUE) #passed the test
}
    else{
      print(FALSE) #failed the test
    }
}
pass_fail(70)
pass_fail(65)
pass_fail(45)
pass_fail(55)
pass_fail
```

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#if else method
pass_fail_result <- pass_fail(70)
pass_fail_result
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?
```{r}
#ifelse method2

pass_fail2 <- function(x){
  ifelse(x>49, print(TRUE), print(FALSE))
}
pass_fail2(70)
pass_fail2(65)
pass_fail2(45)
pass_fail2(55)
pass_fail2
```

> Answer: The 'if' and 'else' method worked and the 'ifelse' didn't quite worked, it has somehow printed the outcome twice in the console. I am not certain why this is happening, but I assume it might be related to the way data is stored and access in 'inelse' is through a vector?  